---
title: "Exploratory Data Analysis of Diabetes Health Indicators"
author: Keshav Ramesh
format: html
---
## Introduction
Here we are going to analyse and explore the 2015 BRFSS Diabetes Health Indicators data set 

**Variables**

Response variable: Diabetes_binary

Demographics Variables: Age, Education, Smoker, Sex, Income

Clinical Variables: BMI, PhysActivity, HighBP, HighChol, CholCheck, Stroke, HeartDiseaseOrAttack, ...



### Purpose of the EDA

For this EDA I will explore Age, Physical Activity, and BMI, as all three variables have strong well established links as risk factors for Diabetes. 

Physical Activity helps regulate blood glucose, and constant sedentary behavior has been linked to increases diabetic risk.  

Age leads to diabetic incidence rising as beta-cell function declines

BMI is a well established risk factor and high amount of fat lead to insulin resistance





```{r}
# Read In packages
library(tidyverse)
```

Read in data
```{r}

data_raw = read_csv("diabetes_binary_health_indicators_BRFSS2015.csv")

# convert variables to factors
diabetes_data = data_raw %>%
  mutate(
    Diabetes_binary = factor(Diabetes_binary, levels = c(0, 1), labels = c("No Diabetes", "Diabetes")),
    PhysActivity = factor(PhysActivity, levels = c(0, 1), labels = c("No", "Yes")),
    AgeGroup = factor(Age, levels = 1:13, labels = c(
  "18–24",
  "25–29",
  "30–34",
  "35–39",
  "40–44",
  "45–49",
  "50–54",
  "55–59",
  "60–64",
  "65–69",
  "70–74",
  "75–79",
  "80+"
))
)

## Check for missingness

summary(diabetes_data)
colSums(is.na(diabetes_data))
```

### Univariate Exploration

Here I am going to check the proportion of diabetics in the dataset

```{r}
diabetes_data %>%
  count(Diabetes_binary) %>%
  mutate(Percent = n / sum(n) * 100)


diabetes_data %>% ggplot(
  aes(
    x = Diabetes_binary, 
    fill = Diabetes_binary)) +
  geom_bar() +
  labs(title = "Diabetes Status Distribution",
       x = "Diabetes",
       y = "Count") +
  theme_minimal()
```

Looking at the age distributions throughout the dataset

```{r}
diabetes_data %>%
  ggplot(aes(x = AgeGroup, fill = AgeGroup)) +
  geom_bar() +
  labs(
    title = "Age Group Distribution",
       x = "Age Group",
       y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")

```


Looking at the physical activity among participants:

```{r}
diabetes_data %>%
  ggplot(aes(x = PhysActivity, fill = PhysActivity)) +
  geom_bar() +
  labs(title = "Physical Activity Distribution",
       x = "Physically Active?",
       y = "Count") +
  theme_minimal()
```

Looking at BMI data across the data

```{r}
diabetes_data %>%
  ggplot(aes(x = BMI)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(title = "BMI Distribution",
       x = "BMI",
       y = "Count") +
  theme_minimal()
```

### Bivariate exploration

See how diabetes changes between age groups 

```{r}
diabetes_data %>%
  ggplot(aes(x = AgeGroup, fill = Diabetes_binary)) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Diabetes by Age Group", x = "Age Group", y = "Proportion") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() 

```


diabetes vs physical activity
```{r}
diabetes_data %>%
  ggplot(aes(x = PhysActivity, fill = Diabetes_binary)) +
  geom_bar(position = "fill") +
  labs(title = "Diabetes by Physical Activity", x = "Physically Active", y = "Proportion") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

```
diabetes vs BMI

```{r}
diabetes_data %>%
  ggplot(aes(x = Diabetes_binary, y = BMI, fill = Diabetes_binary)) +
  geom_violin() +
  labs(title = "BMI by Diabetes Status", x = "Diabetes Status", y = "BMI") +
  theme_minimal()

```

```{r}
diabetes_data %>%
  ggplot(aes(x = BMI, fill = Diabetes_binary)) +
  geom_density(alpha = 0.5) +
  labs(title = "BMI Density by Diabetes Status", x = "BMI", y = "Density") +
  theme_minimal()

```



[Click here for the Modeling Page](Modeling.html)




































