---
title: "Modeling: Predicting Diabetes Risk"
author: "Keshav Ramesh"
format: html
---

## Introduction
Here we are going to analyse and explore the 2015 BRFSS Diabetes Health Indicators data set.

**Variables**

Response variable: Diabetes_binary

Demographics Variables: Age, Education, Smoker, Sex, Income

Clinical Variables: BMI, PhysActivity, HighBP, HighChol, CholCheck, Stroke, HeartDiseaseOrAttack, ...

For this EDA I will explore Age, Physical Activity, and BMI, as all three variables have strong well established links as risk factors for Diabetes. 

Physical Activity helps regulate blood glucose, and constant sedentary behavior has been linked to increases diabetic risk.  

Age leads to diabetic incidence rising as beta-cell function declines

BMI is a well established risk factor and high amount of fat lead to insulin resistance

In addition to these from the exploratory data analysis we will also be including high cholesterol and high blood pressure

High Blood Pressure Frequently occurs alongside diabetes and is considered a common comorbidity. Chronically high blood pressure can damage blood vessels and impact insulin delivery increasing diabetic risk.

High cholesterol is often found in individuals with type 2 diabetes. It can affect underlying insulin resistance and cardiovascular function making it important to consider in predictive modeling.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(caret)
library(yardstick)
library(glmnet)
library(rpart.plot)
library(ranger)
```

```{r}
data_raw = read_csv("diabetes_binary_health_indicators_BRFSS2015.csv")

# convert variables to factors
diabetes_data <- data_raw %>%
  mutate(
    Diabetes_binary = factor(Diabetes_binary, levels = c(0, 1), labels = c("No Diabetes", "Diabetes")),
    PhysActivity = factor(PhysActivity, levels = c(0, 1), labels = c("No", "Yes")),
    HighChol = factor(HighChol, levels = c(0, 1), labels = c("No", "Yes")),
    HighBP = factor(HighBP, levels = c(0, 1), labels = c("No", "Yes")),
    AgeGroup = factor(Age, levels = 1:13, labels = c(
      "18–24",
      "25–29",
      "30–34",
      "35–39",
      "40–44",
      "45–49",
      "50–54",
      "55–59",
      "60–64",
      "65–69",
      "70–74",
      "75–79",
      "80+"
    ))
  )
```

```{r}


set.seed(101)

#Split into testing and training sets
diabetes_split = initial_split(diabetes_data, prop = .7)
diabetes_train = training(diabetes_split)
diabetes_test = testing(diabetes_split)


## create cv folds
cv_folds = vfold_cv(diabetes_train, 5)

```

## Log Regression Model

A logistic regression model is a classification algorithm used to model the probability of a binary outcome. Unlike linear regression which predicts a continuous value, logistic regression estimates the probability given inputs in particular categories. For this data set the response variable is diabetes _binary making logistic regression an appropriate technique.

This model works by transforming linear predictors using a sigmoid function which maps the values and as allows us to interpret the output as probability.


We will use logistic regression for this data set because the target variable is binary, the data set offers appropriate explanatory variables that relate to the probability of having diabetes, and it is a strong model to form a baseline before exploring other multiple alternatives.


```{r}

## Create recipies

diabetes_rec1 = recipe(Diabetes_binary ~ BMI + AgeGroup + PhysActivity, data = diabetes_train) %>%
  step_normalize(all_numeric(), -Diabetes_binary)

diabetes_rec2 = recipe(Diabetes_binary ~ BMI + AgeGroup + PhysActivity + HighBP, data = diabetes_train) %>%
  step_normalize(all_numeric(), -Diabetes_binary)

diabetes_rec3 = recipe(Diabetes_binary ~ BMI + AgeGroup + PhysActivity + HighChol + HighBP, data = diabetes_train) %>%
  step_normalize(all_numeric(), -Diabetes_binary)
```

### Create Workflows
```{r}
# set engine
log_spec = logistic_reg() %>%
  set_engine("glm")

## create workflows
log_workflow1 = workflow() %>%
  add_model(log_spec) %>%
  add_recipe(diabetes_rec1)

log_workflow2 = workflow() %>%
  add_model(log_spec) %>%
  add_recipe(diabetes_rec2)

log_workflow3 = workflow() %>%
  add_model(log_spec) %>%
  add_recipe(diabetes_rec3)


```

### Fit to CV Folds
```{r}
log_res_fit1 = log_workflow1 %>%
  fit_resamples(cv_folds, metrics = metric_set(accuracy,mn_log_loss))

log_res_fit2 = log_workflow2 %>%
  fit_resamples(cv_folds, metrics = metric_set(accuracy,mn_log_loss))

log_res_fit3 = log_workflow3 %>%
  fit_resamples(cv_folds, metrics = metric_set(accuracy,mn_log_loss))
```

### Collect Metrics
```{r}
rbind(log_res_fit1 %>% collect_metrics(),
log_res_fit2 %>% collect_metrics(),
log_res_fit3 %>% collect_metrics()) %>%
mutate(Model = c("Model1", "Model1", "Model2", "Model2", "Model3", "Model3")) %>%
select(Model, everything())
```
Model 3 has the lowest log loss of any of the three potential models therefore we will be using that model moving forward.

### Finding COnfustion Matrix
```{r}
log_train_fit = log_workflow3 %>%
  fit(diabetes_train)

conf_mat(diabetes_train %>%
           mutate(estimate = log_train_fit %>% 
                    predict(diabetes_train) %>%
                    pull()), Diabetes_binary, estimate)

conf_mat(diabetes_test %>%
           mutate(estimate = log_train_fit %>% 
                    predict(diabetes_test) %>%
                    pull()), Diabetes_binary, estimate)


```

### Testing on Test Set
```{r}
log_workflow3 %>%
  last_fit(diabetes_split, metrics = metric_set(mn_log_loss)) %>%
  collect_metrics()

```
### Fit to the entire dataset

```{r}
# fit to the entire dataset
final_log_model = log_workflow3 %>%
  fit(diabetes_data)
#Tidy final model output
tidy(final_log_model)
## test on test set
log_final_fit = log_workflow3 %>%
  last_fit(diabetes_split, metrics = metric_set(accuracy, mn_log_loss))
#collect metrics
log_final_fit %>%
  collect_metrics

log_final_model_train <- log_workflow3 %>%
  fit(data = diabetes_train)
log_final_model_train
```


## Classification Tree

A classification tree is a non paramedic model used for predicting categorical outcomes. It works by splitting the data set based on values of predictor variables to create decision rules that relate to predictions. Each node of the tree represents a decision based off one of the predictors in each terminal node assigns a prediction.

We can use a classification tree for this data because they are very highly interpretable and easy to visualize. They can readily capture non-linear relationships as well as interactions between different variables, and they don't make any assumptions about the distribution of predictors or response variables. However since trees can easily overfit it is important to tune the complexity to balance fit.

### Define model and engine
```{r}

## set model and engine
tree_spec = decision_tree(
  cost_complexity = tune(),
  tree_depth = tune(),
  min_n = 20
) %>%
  set_engine("rpart") %>%
  set_mode("classification")
```

### Create Workflow
```{r}
tree_workflow = workflow() %>%
  add_recipe(diabetes_rec3) %>%
  add_model(tree_spec)

```

### Select tuning parameters

```{r}
## make and tune the model
tree_grid = grid_regular(
  cost_complexity(),
  tree_depth(),
  levels = list(cost_complexity = 10, tree_depth = 5)
)

tree_fits = tree_workflow %>% 
  tune_grid(resamples = cv_folds,
            grid = tree_grid,
            metrics = metric_set(mn_log_loss))
## collect metrics
tree_fits %>%
  collect_metrics() %>%
  filter(.metric == "mn_log_loss") %>%
  arrange(mean)
```

### Select the Best model 
```{r}
# find best model
tree_best_params = select_best(tree_fits, metric = "mn_log_loss")

#final workflow
tree_final_workflow = tree_workflow %>%
  finalize_workflow(tree_best_params)
# test on test dataset
tree_final_fit = tree_final_workflow %>%
  last_fit(diabetes_split, metrics = metric_set(accuracy, mn_log_loss))
## collect metrics
tree_final_fit %>%
  collect_metrics()

tree_final_model_train <- tree_final_workflow %>%
  fit(data = diabetes_train)

tree_final_model_train

tree_final_model = extract_workflow(tree_final_fit) 

# Plot the classification tree
tree_final_model %>%
  extract_fit_engine() %>%
  rpart.plot::rpart.plot(roundint = FALSE)

```



## Random Forest 

A random forest is an ensemble learning method that builds a collection of decision trees and combines their predictions to produce more accurate and robust models. Each tree is trained on a bootstrap sample of the data and each split is a random subset of predictors rather than all of them. This randomness helps reduce variance and makes the final model less prone to overfitting.

We will use a random forest because while classification trees very intuitive they're also highly variable models and small changes in the data can lead to wildly different trees. Random forests addressed this limitation by averaging the predictions, and lead to better generalizations.

### Define Model

```{r}

## set model and engine
rf_spec = rand_forest(mtry = tune()) %>% 
  set_engine("ranger") %>% 
  set_mode("classification")
```

### Create Workflows
```{r}

rf_workflow = workflow() %>% 
  add_recipe(diabetes_rec3) %>%
  add_model(rf_spec)
```

### Tuning Model

```{r}
rf_grid <- grid_regular(mtry(range = c(1, 4)), levels = 5)

# tune model
rf_fit = rf_workflow %>%
  tune_grid(
    resamples = cv_folds,
            grid = rf_grid,
            metrics = metric_set(accuracy, mn_log_loss))

rf_fit %>% 
  collect_metrics() %>% 
  filter(.metric == "mn_log_loss") %>% 
  arrange(mean)
```

```{r}

## select best model
rf_best_params = select_best(rf_fit, metric = "mn_log_loss")
rf_best_params

#final workflow
rf_final_workflow = rf_workflow %>%
  finalize_workflow(rf_best_params)

##test on test set
rf_final_fit = rf_final_workflow %>% 
  last_fit(diabetes_split, metrics = metric_set(accuracy, mn_log_loss))

## collect metrics
rf_final_fit %>%
  collect_metrics()

rf_final_model_train <- rf_final_workflow %>%
  fit(data = diabetes_train)

rf_final_model_train

```

## Comparing Models

```{r}

log_final_fit %>%
  collect_metrics

tree_final_fit %>%
  collect_metrics()

rf_final_fit %>%
  collect_metrics
```
Looking at the log loss comparison of all of the models the random forest has the lowest log loss.




[Back to EDA Page](EDA.html)


