---
title: "Modeling: Predicting Diabetes Risk"
author: "Keshav Ramesh"
format: html
---

## Introduction


```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(caret)
library(yardstick)
library(glmnet)
library(rpart.plot)
library(ranger)
```

```{r}
data_raw = read_csv("diabetes_binary_health_indicators_BRFSS2015.csv")

# convert variables to factors
diabetes_data <- data_raw %>%
  mutate(
    Diabetes_binary = factor(Diabetes_binary, levels = c(0, 1), labels = c("No Diabetes", "Diabetes")),
    PhysActivity = factor(PhysActivity, levels = c(0, 1), labels = c("No", "Yes")),
    HighChol = factor(HighChol, levels = c(0, 1), labels = c("No", "Yes")),
    HighBP = factor(HighBP, levels = c(0, 1), labels = c("No", "Yes")),
    AgeGroup = factor(Age, levels = 1:13, labels = c(
      "18–24",
      "25–29",
      "30–34",
      "35–39",
      "40–44",
      "45–49",
      "50–54",
      "55–59",
      "60–64",
      "65–69",
      "70–74",
      "75–79",
      "80+"
    ))
  )
```

```{r}
set.seed(101)

diabetes_split = initial_split(diabetes_data, prop = .7)
diabetes_train = training(diabetes_split)
diabetes_test = testing(diabetes_split)

cv_folds = vfold_cv(diabetes_train, 5)

```

## Log Regression Model

```{r}

diabetes_rec1 = recipe(Diabetes_binary ~ BMI + AgeGroup + PhysActivity, data = diabetes_train) %>%
  step_normalize(all_numeric(), -Diabetes_binary)

diabetes_rec2 = recipe(Diabetes_binary ~ BMI + AgeGroup + PhysActivity + HighBP, data = diabetes_train) %>%
  step_normalize(all_numeric(), -Diabetes_binary)

diabetes_rec3 = recipe(Diabetes_binary ~ BMI + AgeGroup + PhysActivity + HighChol + HighBP, data = diabetes_train) %>%
  step_normalize(all_numeric(), -Diabetes_binary)
```

### Create Workflows
```{r}

log_spec = logistic_reg() %>%
  set_engine("glm")

log_workflow1 = workflow() %>%
  add_model(log_spec) %>%
  add_recipe(diabetes_rec1)

log_workflow2 = workflow() %>%
  add_model(log_spec) %>%
  add_recipe(diabetes_rec2)

log_workflow3 = workflow() %>%
  add_model(log_spec) %>%
  add_recipe(diabetes_rec3)


```

### Fit to CV Folds
```{r}
log_res_fit1 = log_workflow1 %>%
  fit_resamples(cv_folds, metrics = metric_set(accuracy,mn_log_loss))

log_res_fit2 = log_workflow2 %>%
  fit_resamples(cv_folds, metrics = metric_set(accuracy,mn_log_loss))

log_res_fit3 = log_workflow3 %>%
  fit_resamples(cv_folds, metrics = metric_set(accuracy,mn_log_loss))
```

### Collect Metrics
```{r}
rbind(log_res_fit1 %>% collect_metrics(),
log_res_fit2 %>% collect_metrics(),
log_res_fit3 %>% collect_metrics()) %>%
mutate(Model = c("Model1", "Model1", "Model2", "Model2", "Model3", "Model3")) %>%
select(Model, everything())
```

### Finding COnfustion Matrix
```{r}
log_train_fit = log_workflow3 %>%
  fit(diabetes_train)

conf_mat(diabetes_train %>%
           mutate(estimate = log_train_fit %>% 
                    predict(diabetes_train) %>%
                    pull()), Diabetes_binary, estimate)

conf_mat(diabetes_test %>%
           mutate(estimate = log_train_fit %>% 
                    predict(diabetes_test) %>%
                    pull()), Diabetes_binary, estimate)


```

### Testing on Test Set
```{r}
log_workflow3 %>%
  last_fit(diabetes_split, metrics = metric_set(mn_log_loss)) %>%
  collect_metrics()

```
### Fit to the entire dataset

```{r}
final_log_model = log_workflow3 %>%
  fit(diabetes_data)

tidy(final_log_model)

log_final_fit = log_workflow3 %>%
  last_fit(diabetes_split, metrics = metric_set(accuracy, mn_log_loss))

log_final_fit %>%
  collect_metrics
```


## Classification Tree

### Define model and engine
```{r}
tree_spec = decision_tree(
  cost_complexity = tune(),
  tree_depth = tune(),
  min_n = 20
) %>%
  set_engine("rpart") %>%
  set_mode("classification")
```

### Create Workflow
```{r}
tree_workflow = workflow() %>%
  add_recipe(diabetes_rec3) %>%
  add_model(tree_spec)

```

### Select tuning parameters

```{r}

tree_grid = grid_regular(
  cost_complexity(),
  tree_depth(),
  levels = list(cost_complexity = 10, tree_depth = 5)
)

tree_fits = tree_workflow %>% 
  tune_grid(resamples = cv_folds,
            grid = tree_grid,
            metrics = metric_set(mn_log_loss))

tree_fits %>%
  collect_metrics() %>%
  filter(.metric == "mn_log_loss") %>%
  arrange(mean)
```

### Select the Best model 
```{r}

tree_best_params = select_best(tree_fits, metric = "mn_log_loss")

tree_final_workflow = tree_workflow %>%
  finalize_workflow(tree_best_params)

tree_final_fit = tree_final_workflow %>%
  last_fit(diabetes_split, metrics = metric_set(accuracy, mn_log_loss))

tree_final_fit %>%
  collect_metrics()

tree_final_model = extract_workflow(tree_final_fit) 

tree_final_model %>%
  extract_fit_engine() %>%
  rpart.plot::rpart.plot(roundint = FALSE)

```



## Random Forest 

### Define Model

```{r}
rf_spec = rand_forest(mtry = tune()) %>% 
  set_engine("ranger") %>% 
  set_mode("classification")
```

### Create Workflows
```{r}
rf_workflow = workflow() %>% 
  add_recipe(diabetes_rec3) %>%
  add_model(rf_spec)
```

### Fit to CV folds

```{r}
rf_grid <- grid_regular(mtry(range = c(1, 4)), levels = 5)

rf_fit = rf_workflow %>%
  tune_grid(
    resamples = cv_folds,
            grid = rf_grid,
            metrics = metric_set(accuracy, mn_log_loss))

rf_fit %>% 
  collect_metrics() %>% 
  filter(.metric == "mn_log_loss") %>% 
  arrange(mean)
```

```{r}
rf_best_params = select_best(rf_fit, metric = "mn_log_loss")
rf_best_params


rf_final_workflow = rf_workflow %>%
finalize_workflow(rf_best_params)

rf_final_fit = rf_final_workflow %>%
last_fit(diabetes_split, metrics = metric_set(accuracy, mn_log_loss))

rf_final_fit %>%
  collect_metrics()
```

## Comparting Models

```{r}

log_final_fit %>%
  collect_metrics

tree_final_fit %>%
  collect_metrics()

rf_final_fit %>%
  collect_metrics
```

Best model is Random forest 







